---
title: "EmptyDroplet_doublet_ambientRNA_detection"
author: "JD Luo"
date: "2023-07-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Process counting with Cellranger

## Gathering sample information
- Assign path to the cellranger counting matrix of each sample
- rawMTX: the path to raw count matrix
- filtMTX: the path to filtered count matrix 
- time: time point
- rep: replications
```{r samInfo,include=TRUE,eval=FALSE}
sample_sheet <- data.frame(sample_id=c("P1_rep1","P1_rep2","P14_rep1","P14_rep2",
                                       "Ad2wk_rep1","Ad2wk_rep2","Ad2mon_rep1","Ad2mon_rep2"),
                           rawMTX=c("P1_rep1/raw_feature_bc_matrix/",
                                    "P1_rep2/raw_feature_bc_matrix/",
                                    "P14_rep1/raw_feature_bc_matrix/",
                                    "P14_rep2/raw_feature_bc_matrix/",
                                    "2wk_rep1/raw_feature_bc_matrix/",
                                    "2wk_rep2/raw_feature_bc_matrix/",
                                    "2mon_rep1/raw_feature_bc_matrix/",
                                    "2mon_rep2/raw_feature_bc_matrix/"),
                           filtMTX=c("P1_rep1/filtered_feature_bc_matrix/",
                                     "P1_rep2/filtered_feature_bc_matrix/",
                                     "P14_rep1/filtered_feature_bc_matrix/",
                                     "P14_rep2/filtered_feature_bc_matrix/",
                                     "2wk_rep1/filtered_feature_bc_matrix/",
                                     "2wk_rep2/filtered_feature_bc_matrix/",
                                     "2mon_rep1/filtered_feature_bc_matrix/",
                                     "2mon_rep2/filtered_feature_bc_matrix/"),
                           time=c(rep("P1",2),rep("P14",2),rep("Ad2wk",2),rep("Ad2mon",2)),
                           rep=c(rep(c("rep1","rep2"),4)),
                           stringsAsFactors = FALSE)
write.table(sample_sheet,file = "sampleSheet_snSeq_AntNeuron_allSample_20230227.csv",
            sep=",",quote = FALSE,row.names = FALSE)
```

## Detect and remove empty droplets
* The empty droplets of each sample were identified and removed by using DropletUtils. [link to bioconductor page](https://bioconductor.org/packages/release/bioc/html/DropletUtils.html)
* Input: raw count matrix generated by Cellranger
* Steps:
  + read raw count matrix into SingleCellExperiment object (sce.pbmc): DropletUtils::read10XCounts()
  + detect empty droplets by the distribution of UMI count per cell: DropletUtils::emptyDrops()
  + subset the cells whose FDR < 0.0001
  + extract the count matrix from subset SingleCellExperiment object
* Output in a list of matrix format
  + raw count matrix extracted from cellranger
  + count matrix filtered by using DropletUtls
```{r emptyDroplet_proc,include=TRUE,eval=FALSE}
# read sample sheet containing raw/filt matrix of cellranger for each sample
sample_sheet <- read.delim("sampleSheet_snSeq_AntNeuron_allSample_20230227.csv",
                           sep=",",stringsAsFactors = FALSE)
#
# process raw matrix, detect empty droplets, and remove empty droplets
mtx_res <- lapply(sample_sheet$sample_id,function(samID,samSheet){
  raw_mtx <- samSheet$rawMTX[samSheet$sample_id==samID] # assign the path to raw count matrix (GEX format)
  sce.pbmc <- read10xCounts(raw_mtx, col.names=TRUE) # read raw count matrix file into SingleCellExperiment object
  # raw_mat <- counts(sce.pbmc) # extact count matix in matrix format
  #
  # for knee plot ~ redundant
  # bcrank <- barcodeRanks(counts(sce.pbmc)) 
  # knee <- metadata(bcrank)$knee
  # cell_kp <- rownames(bcrank)[bcrank$total >= knee]
  #
  set.seed(100)
  e.out <- emptyDrops(counts(sce.pbmc)) # detect empty droplets by the distribution
  # See ?emptyDrops for an explanation of why there are NA values.
  # summary(e.out$FDR <= 0.001)
  # table(Sig=e.out$FDR <= 0.001, Limited=e.out$Limited)
  #
  sce.sub <- sce.pbmc[,which(e.out$FDR < 0.0001)]
  filt_mat <- counts(sce.sub)
  #
  res <- list("raw"=raw_mat,"filt"=filt_mat)
  base::save("raw_mat","filt_mat",file = paste0("procMTX_",samID,"_20230227.RData"))
  # abc <- base::load("procMTX_P1_rep1_20230227.RData")
  return(res)},sample_sheet)
names(mtx_res) <- sample_sheet$sample_id
```

## Detect doublets
* Doublets were detected with Scrublet [link](https://github.com/swolock/scrublet).
* Input: count matrix after removing empty droplets with DropletUtils.
* Steps:
  + Invoke scrublet into R with reticulate
  + Detect doublets with scrubelt
  + Transfer doublet information into Seurat object
* Output: Seurat object with doublet information
```{r doublet_det,include=TRUE,eval=FALSE}
## Load filtered matrix into Seurat object
# Assign the files of DropletUtil-processed matrix 
mtx_files <- setNames(c("procMTX_P1_rep1_20230227.RData","procMTX_P1_rep2_20230227.RData",
                        "procMTX_P14_rep1_20230227.RData","procMTX_P14_rep2_20230227.RData",
                        "procMTX_Ad2wk_rep1_20230227.RData","procMTX_Ad2wk_rep2_20230227.RData",
                        "procMTX_Ad2mon_rep1_20230227.RData","procMTX_Ad2mon_rep2_20230227.RData"),
                      c("P1_rep1","P1_rep2","P14_rep1","P14_rep2","Ad2wk_rep1","Ad2wk_rep2","Ad2mon_rep1","Ad2mon_rep2"))
#
# Load raw matrix
mtxRes_raw <- lapply(mtx_files,function(mtx){base::load(mtx);return(raw_mat)});names(mtxRes_raw) <- names(mtx_files)
#
# Load filtered matrix
mtxRes_filt <- lapply(mtx_files,function(mtx){base::load(mtx);return(filt_mat)});names(mtxRes_filt) <- names(mtx_files)
##
## Load filtered matrix into Seurat object
seu_list <- lapply(names(mtxRes_filt),function(samID,mtxRes_filt){
  mtx <- mtxRes_filt[[samID]]
  seu_obj <- Seurat::CreateSeuratObject(mtx)
  seu_obj[["dset"]] <- samID
  seu_obj[["percent.mt"]] <- PercentageFeatureSet(seu_obj,pattern = "^MT-")
  seu_obj <- SCTransform(seu_obj,verbose = FALSE)},mtxRes_filt)
names(seu_list) <- names(mtxRes_filt)
##
## Detect doublets with scrublet
doublet_eval <- function(seu_obj,py_path=NULL){
  require(reticulate)
  require(Seurat)
  #
  Sys.setenv('RETICULATE_PYTHON'=py_path)
  scr <- import("scrublet")
  scrub <- scr$Scrublet(t(as.matrix(Seurat::GetAssayData(seu_obj,assay="RNA",slot = "counts"))))
  doublet <- scrub$scrub_doublets()
  names(doublet) <- c("doublet_score","doublet")
  tab <- data.frame(cell_id=Cells(seu_obj),
                    doublet_score=doublet$doublet_score,
                    doublet=doublet$doublet,stringsAsFactors = FALSE)
  return(tab)}
doublet_list <- lapply(seu_list,doublet_eval,
                       py_path="/Users/JDLuo/opt/anaconda3/envs/reticulate/bin/python")
##
## Transfer doublet information into Seurat object
sample_id <- names(seu_list)
seu_list <- lapply(names(seu_list),function(samID,seu_list,doublet_list){
  seu_obj <- seu_list[[samID]]
  doublet_dat <- doublet_list[[samID]]
  seu_obj[["doubletScore"]] <- doublet_dat$doublet_score
  seu_obj[["doublet"]] <- doublet_dat$doublet
  return(seu_obj)},seu_list,doublet_list)
names(seu_list) <- sample_id
save("seu_list",file = "seuOBJ_antNeuron_oriList.RData")
save("doublet_list",file = "doubletListantNeuron_oriList.RData")
##
## END
```

## Remove ambient RNA
* Ambient RNA fractions were estimated and corrected with SoupX [link](https://github.com/constantAmateur/SoupX).
* Input:
  + raw matrix generated by cellranger 
  + filterd matrix generated by DropletUtils
  + clustering of filtered matrix
* Steps:
  + load raw/filtered matrix into SoupX channel: SoupX::SoupChannel()
  + set cluster information: SoupX::setClusters()
  + estimate contamination fractions automatically: autoEstCont()
    Set *forceAccept=TRUE* if the ambient RNA contamination rate would be higher than 0.2
  + correct counts: SouopX::adjustCounts()
  + Load corrected matrix into Seurat object
* output: Seurat object after ambient RNA correction
* import doublet information into Seurat objects
```{r rmAmb,include=TRUE,eval=FALSE}
## load raw matrix
mtx_files <- setNames(c("procMTX_P1_rep1_20230227.RData","procMTX_P1_rep2_20230227.RData",
                        "procMTX_P14_rep1_20230227.RData","procMTX_P14_rep2_20230227.RData",
                        "procMTX_Ad2wk_rep1_20230227.RData","procMTX_Ad2wk_rep2_20230227.RData",
                        "procMTX_Ad2mon_rep1_20230227.RData","procMTX_Ad2mon_rep2_20230227.RData"),
                      c("P1_rep1","P1_rep2","P14_rep1","P14_rep2","Ad2wk_rep1","Ad2wk_rep2","Ad2mon_rep1","Ad2mon_rep2"))
#
# Load raw matrix
mtxRes_raw <- lapply(mtx_files,function(mtx){base::load(mtx);return(raw_mat)});names(mtxRes_raw) <- names(mtx_files)
#
# Load filtered matrix
mtxRes_filt <- lapply(mtx_files,function(mtx){base::load(mtx);return(filt_mat)});names(mtxRes_filt) <- names(mtx_files)
##
## make clustering with default setting
load("seuOBJ_antNeuron_oriList.RData")
quick_clust <- function(seu_obj){
  set.seed(0)
  seu_obj <- Seurat::RunPCA(seu_obj, npcs = 30, verbose = FALSE)
  seu_obj <- Seurat::FindNeighbors(seu_obj, reduction = "pca", dims = 1:10,verbose=FALSE)
  seu_obj <- Seurat::RunUMAP(seu_obj, reduction = "pca", dims = 1:10,verbose=FALSE)
  seu_obj <- Seurat::FindClusters(seu_obj,resolution = 0.5,verbose=FALSE)
  return(seu_obj)}
seu_list <- lapply(seu_list,quick_clust)
clust_list <- lapply(seu_list,function(x){clust <- x$seurat_clusters;return(clust)}) # for the following
##
## remove ambient RNA with SoupX
scLoad_withCorr <- function(samID,clust_list,mtxRes_raw,mtxRes_filt){
  require(SoupX)
  require(Seurat)
  #
  clust <- clust_list[[samID]]
  raw_mtx <- mtxRes_raw[[samID]]
  filt_mtx <- mtxRes_filt[[samID]]
  #
  # SoupX processing
  soupOBJ <- SoupChannel(raw_mtx, filt_mtx)
  soupOBJ <- setClusters(soupOBJ,clust)
  soupOBJ <- autoEstCont(soupOBJ,forceAccept = TRUE)
  soupMTX <- adjustCounts(soupOBJ)
  #
  # Load corrected matrix into Seurat
  seu_obj <- CreateSeuratObject(soupMTX,project=samID)
  seu_obj[["dset"]] <- samID
  seu_obj[["percent.mt"]] <- PercentageFeatureSet(seu_obj,pattern = "^MT-")
  seu_obj <- SCTransform(seu_obj,verbose = FALSE)
  return(seu_obj)}
seu_list <- lapply(names(mtxRes_raw),scLoad_withCorr,clust_list,mtxRes_raw,mtxRes_filt)
names(seu_list) <- names(mtxRes_raw)
##
## Transfer doublet information into Seurat object
load("doubletListantNeuron_oriList.RData")
sample_id <- names(seu_list)
seu_list <- lapply(names(seu_list),function(samID,seu_list,doublet_list){
  seu_obj <- seu_list[[samID]]
  doublet_dat <- doublet_list[[samID]]
  seu_obj[["doubletScore"]] <- doublet_dat$doublet_score
  seu_obj[["doublet"]] <- doublet_dat$doublet
  return(seu_obj)},seu_list,doublet_list)
names(seu_list) <- sample_id
##
## END
```

## Sessions
```{r session,include=TRUE,eval=TRUE}
sessionInfo()
```

Bioinformatics Resource Center, The Rockefeller University